Sequential algorithms to split and merge ultra-high resolution 3D images

### 2017-07-29 paper.tex
Wrote some discussion. Fixed references

\author{Final ordering
  to be discussed when the paper: Yongping/Val\'erie,
  Yuhong Yan, Tristan.\\
  Department of Computer Science and Software Engineering, Concordia University, Montreal, Quebec, Canada
}

@@

\begin{abstract}
  \todo{The abstract goes here. DO NOT USE SPECIAL CHARACTERS,
    SYMBOLS, OR MATH IN YOUR TITLE OR ABSTRACT.} 
\end{abstract}

@@

Three-dimensional images that exceed typical memory size are
increasingly found in a variety of disciplines. BigBrain, for
instance, is a 3D histological image of the human brain that
represents 1~TB of data organized in 3600 planes at full resolution
(1$\mu$m in-plane), and 76~GB at a 40$\mu$m isotropic resolution
commonly used in neurosciences~\cite{amunts2013bigbrain}. Other
examples found in medical imaging, our primary domain of interest,
include \todo{electromicroscopy,spectroscopy-francoise-peyrin}. As
such images would typically be processed on a computing cluster,
possibly using locality-aware file systems such as HDFS, software
libraries are needed to split and merge such images efficiently. We
introduce and compare a family of algorithms to address this problem
in a sequential environment made of a single storage disk. We are
aware of the need for parallel algorithms that would split and merge
images to and from an array of disks but we plan to describe such
algorithms in a subsequent paper based on the results of the present
one. \todo{
There is BigBrain-1 rescanned at 1 micron in the plane. So resolution
is 1x1x20, 7404 slices. If raw data are about 1Tb, then multiply by
100 for an estimation of the size of the new image. }

@@

The literature looks remarkably scarce on the problem described
above. Parallel processing of 3D images has obviously been extensively
studied, but methods have focused more on geometrical approaches to
partition images and on load-balancing and task scheduling techniques
while we aim at algorithms to efficiently split or merge images
regardless of the geometry of the chunks \todo{do a serious literature
review on this topic}. Perhaps this lack of related works could be
attributed to the fact that disk I/O weren't a bottleneck until image
resolution reached a certain threshold. A relevant reference is the
description of the Open Connectome Data Cluster~\cite{burns2013open},
a data warehouse system that allows users to retrieve specific chunks
of large datasets. In this work, the problem of long seek times is
addressed by designing a specific file format based on space-filling
curves, that elegantly preserves spatial proximity on disk. However,
addressing the problem using a specific data organization optimizes
the system for specific chunk geometries, volumetric in this case,
while we aim at algorithms that support multiple types of geometries.

This paper makes the following contributions:
\begin{itemize}
  \item We propose a set of algorithms to split and merge 3D images
    that do not fit in memory.
  \item We model the performance (execution time) of these algorithms,
    based on a simple but realistic disk I/O model.
  \item We evaluate our model and algorithms on a representative
    use-case involving a 1~TB image a hard drive and a solid-state
    drive.
  \item We compare the algorithms in a variety of conditions based on the validated model.
\end{itemize}
Section~\ref{sec:algos} presents our algorithms and their complexity,
Section~\ref{sec:results} details the results and
Sections~\ref{sec:discussion} discusses them.

@@

\subsection{Implementation}

@@

We run experiments using a Dell Precision Tower 3620 with CentOS Linux release
7.3.1611, 32~GB of RAM and two disks:
\begin{itemize}
  \item Hard disk drive (HDD): HGST Travelstar 7K1000, 7200~rpm, 931GiB (1TB), firmware version JB0OA3W0.
    \item Solid-state drive (SSD): SanDisk X400 2.5, 238GiB (256GB),
      firmware version X4130012.
\end{itemize}
Both drives used 512-byte logical sectors, 4096-byte physical sectors,
SATA >3.1 (6.0 Gb/s) and were accessed through the XFS file system.
%smartctl --xall /dev/sda

@@

What is the best algorithm (between 3 and 4), in which context. Did we
solve the seek problem or is there any issue remaining?

File formats:
\begin{itemize}
\item would MINC~\cite{vincent2016minc} support such algorithms? See on-going email discussion
with P. Bellec.
\item some formats optimize the storage for particular split shapes. See
  ndstore~\cite{burns2013open}. Such formats would behave equally
  poorly with chunks that do not comply to this geometry. Our
  algorithms would work equally well with such formats.
\end{itemize}

The performance model might be dependent on the particular
infrastructure (disks, OS and filesystem) used in the experiments. In
particular, caching might behave differently in a different setup.

Reconstructing (therefore splitting) a complete 3D image may not
always be necessary. Re-splitting, for instance, could be done from
existing chunks.

\section{Conclusion}

Future work:
- obliques
- parallel algorithms
- re-splitting
- model

>>>>

\author{Final ordering
  to be discussed: Yongping/Val\'erie,
  Yuhong Yan, Tristan.\\
  Department of Computer Science and Software Engineering, Concordia University, Montreal, Quebec, Canada
}

@@

\begin{abstract}
  Splitting and merging data is a requirement for most parallel
    or distributed processing operations, however, naive algorithms
    perform very poorly on ultra-high resolution 3D images due to
    important seek times. To address this issue, we created two
    algorithms based on in-memory reconstruction of image segments and
    analyzed their complexity in terms of number of seeks. Evaluation
    on a 3000x3000x3000 image of the brain shows that both our
    algorithms can split or merge blocks with a performance similar to
    slices, which solves our initial problem. We conclude that
    splitting and merging large 3D images can be done efficiently
    based on simple data formats.
\end{abstract}

@@

Three-dimensional images that exceed typical memory size are
increasingly found in a variety of disciplines. BigBrain, for
instance, is a 3D histological image of the human brain that
represents 1~TB of data organized in 3600 planes at full resolution
(1$\mu$m in-plane), and 76~GB at a 40$\mu$m isotropic resolution
commonly used in neurosciences~\cite{amunts2013bigbrain}. Other
examples found in medical imaging, our primary domain of interest,
include high-resolution 3D electron microscopy (see, e.g.,
~\cite{bock2011network}) or micro and
nano-tomography~\cite{10.1371/journal.pone.0035691}. As such images
would typically be processed on a computing cluster, possibly using
locality-aware file systems such as HDFS, software libraries are
needed to split and merge such images efficiently. We introduce and
compare a family of algorithms to address this problem in a sequential
environment made of a single storage disk. We are aware of the need
for parallel algorithms that would split and merge images to and from
an array of disks but we plan to describe such algorithms in a
subsequent paper based on the results of the present one. \todo{ There
  is BigBrain-1 rescanned at 1 micron in the plane. So resolution is
  1x1x20, 7404 slices. If raw data are about 1Tb, then multiply by 100
  for an estimation of the size of the new image. }

@@

The literature is remarkably scarce on the problem described
above. Parallel processing of 3D images has obviously been extensively
studied, but methods have focused more on geometrical approaches to
partition images and on load-balancing and task scheduling techniques
while we aim at algorithms to efficiently split or merge images
regardless of the geometry of the chunks \todo{do a serious literature
review on this topic}. Perhaps this lack of related works could be
attributed to the fact that disk I/O weren't a bottleneck until image
resolution reached a certain threshold. A relevant reference is the
description of the Open Connectome Data Cluster~\cite{burns2013open},
a data warehouse system that allows users to retrieve specific chunks
of large datasets. In this work, the problem of long seek times is
addressed by designing a specific file format based on space-filling
curves, that elegantly preserves spatial proximity on disk. However,
addressing the problem using a specific data organization optimizes
the system for specific chunk geometries, volumetric in this case,
while we aim at algorithms that support multiple types of geometries.

This paper makes the following contributions:
\begin{itemize}
  \item We propose a set of algorithms to split and merge 3D images
    that do not fit in memory.
  \item We model the performance (number of seeks) of these algorithms,
    based on a simple but realistic disk I/O model.
  \item We evaluate our model and algorithms on a representative
    use-case involving a 1~TB image a hard drive and a solid-state
    drive.
\end{itemize}
Section~\ref{sec:algos} presents our algorithms and their complexity,
Section~\ref{sec:results} details the results and
Sections~\ref{sec:discussion} discusses them.

@@

\subsection{Implementation}
\label{sec:implementation}

@@

We run experiments using a Dell Precision Tower 3620 with CentOS Linux release
7.3.1611, 32~GB of RAM and two disks:
\begin{itemize}
  \item Hard disk drive (HDD): HGST Travelstar 7K1000, 7200~rpm, 931GiB (1TB), firmware version JB0OA3W0.
    \item Solid-state drive (SSD): SanDisk X400 2.5, 238GiB (256GB),
      firmware version X4130012.
\end{itemize}
Both drives used 512-byte logical sectors, 4096-byte physical sectors,
SATA >3.1 (6.0 Gb/s) and were accessed through the XFS file system
v4.5.0.
%smartctl --xall /dev/sda

\subsection{Execution conditions}

We used release xxx of our library.

\todo{Yongping: mention the cache cleaning thing that you run before every experiment.}

Experiment scripts are available at xxx.

@@

% Problem solved!
Cluster reads and multiple reads are able to reduce to a negligible
amount the overall seek time required to split or merge 3D blocks in a
high-resolution image where data is stored linearly. Both algorithms
performed equivalently on our benchmark and compared to the reference
configuration where slices are merged or split without seeking. Our
initial problem is solved.

% Parallel algorithms
In a real system, high-resolution images are likely to be processed on
computing clusters, for instance using software from the Hadoop
project, in particular the Hadoop Distributed File
System~\cite{shvachko2010hadoop}. In this context parallel split-and-merge algorithms
will be required, since the various splits of a large image would be
uploaded to different disks concurrently. In the same vein,
``re-spliting'' algorithms would have to be devised in case a split
image need to be split in a different geometry. Cluster reads and
multiple reads could be used as starting points for such algorithms.

% Several other factors might influence performance
I/O optimization is a holistic problem that is in practice highly
dependent on the hardware used, firmware, operating system, file
system and programming language. Caching occurs at various levels and
might always influence performance, potentially differently depending
on the split or merge algorithm used. In some disks, seek time greatly
varies with the seek distance, which would open the door to additional
opportunities for I/O optimization. Interactions between those
components might also result in performance differences particular to
a particular algorithm. To ensure the portability of our library
across systems and configurations, we focused on reducing the overall
number of seeks and ignored specific system configurations. We
demonstrated the performance of our methods on both an HDD and an SSD
disk, using state-of-the-art and widely used versions of Linux
(CentOS7) and file systems (XFS v4.5.0).

% File formats
In terms of file formats, our results demonstrate that simple imaging
formats might be used for the split-and-merge problem without
performance loss compared to more complex formats that try to preserve
spatial locality on disk, for instance MINC 2.0~\cite{vincent2016minc}
or the format based on space-filling curves mentioned
in~\cite{burns2013open}. This is of major interest in the current
open-science context since simpler formats favor data-sharing and
interoperability.  Moreover, our algorithms could potentially be
adapted to any split geometry, even though we demonstrated them on
slices and blocks only, while file formats inevitably assume a
particular geometry. For instance, the format in~\cite{burns2013open}
wouldn't perform nicely with a naive splitting algorithm if slices had
to be extracted instead of blocks.  Two points need to be made to be
fair though. First, we aimed at extracting all the blocks while the
problem targeted by the format in~\cite{burns2013open} was to extract
a single block from a large image. Second, we have not considered
on-the-fly data compression yet, which is a unique feature of MINC
2.0.
